---
title: "Bibliographic analysis of occupational health journals"
author: 
  - "Javier Mancilla Galindo"
date: today
execute: 
  echo: false
  warning: false
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
    code-links:
        - text: "GitHub"
          href: https://github.com/UtrechtUniversity/DOA-journal-occupational
          icon: github
  pdf: 
    toc: true
    documentclass: scrartcl
editor: source
---

```{r}
#| include: false  

# Create directories for sub-folders  
inputfolder <- "../data/raw"
psfolder <- "../data/processed"
tempfolder <- "../data/temp"
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"

dir.create(inputfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
dir.create(tempfolder, showWarnings = FALSE)
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
```

```{r}

if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  writexl,
  readxl,
  rvest,
  rcrossref,
  httr,
  jsonlite,
  gt,
  report
)

```

The list of journals indexed for OSHLINE® was extracted on 23/04/2025 from the Canadian Centre for Occupational Health and Safety [public website](https://www.ccohs.ca/products/supplements/oshline/oshline_journal_list.html). This list contains journal names and their ISSN.

```{r}
#| echo: true 
OSHLINE_journals <- read_html(paste0(inputfolder,"/oshline_journal_list.html"))
```

A dataset will be generated from the html file by using functions from the [`rvest`](https://rvest.tidyverse.org/) package to obtain a column with the journal name and the ISSN.

```{r}
# Extract and clean text from html file
journal_text <- OSHLINE_journals %>%
  html_elements("div.centralContainer p") %>%
  html_text(trim = TRUE)

# Drop first 3 non-journal entries
journal_text <- journal_text[-c(1:3)]

# Split entries that contain multiple journals (multiple ISSNs in one line)
# Use regex to split lines where ISSN pattern is followed by another ISSN
split_lines <- unlist(strsplit(journal_text, "(?<=\\d{4}-\\d{3}[\\dX])(?=\\S)", perl = TRUE))

# Clean whitespace and remove empties
split_lines <- str_trim(split_lines)
split_lines <- split_lines[split_lines != ""]

# Identify correct ISSNs, filter valid lines
journal_df <- tibble(raw_entry = split_lines) %>%
  mutate(
    issn = str_extract_all(raw_entry, "\\d{4}-\\d{3}[\\dX]"),
    count = map_int(issn, length)
  ) %>%
  unnest(issn) %>%
  mutate(journal_name = str_trim(str_remove(raw_entry, issn))) %>%
  select(journal_name, issn, raw_entry)

# Examine
glimpse(journal_df)

# Save as excel file to resolve some remaining collapsed rows and errors
write_xlsx(journal_df, path = file.path(tempfolder, "OSHLINE_journals.xlsx")) 
```

After manual inspection and checks for errors, two journals were removed as the ISSN was not registered for one (*Facility Safety Management*) and the correct ISSN could not be identified for a journal with no exact name matches (*Health Promotion*). *Environmental Carcinogenesis and Ecotoxicology Reviews* (ISSN = 1059-0501) was removed as this was a duplicate of the *Journal of Environmental Science and Health. Part C, Environmental Carcinogenesis & Ecotoxicology Reviews* (ISSN = 1059-0501).

```{r}
#| echo: true 
journal_df <- read_excel(path = file.path(psfolder,"OSHLINE_journals.xlsx"))
```

```{r}
#| include: false 
# Check if all have unique ISSN 
journal_df %>%
  group_by(issn) %>%
  filter(n() > 1)
```

This resulted in a total of **n = `r count(journal_df)`** journals.

The metadata for these journals will be enriched by using the [`rcrossref`](https://rvest.tidyverse.org/) package by first searching for the ISSN. For cases where there is no ISSN match, journal titles will be searched. 

```{r}
#| echo: true 

source("scripts/crossref_metadata.R")
```

```{r}
#| echo: true 
journals_crossref <- read_excel(path = file.path(psfolder,"OSHLINE_crossref.xlsx"))
```

Retrieve electronic issn from ISSN portal 
```{r}

get_eissn_simple <- function(issn) {
  # Skip NA values
  if (is.na(issn)) return(NA_character_)
  
  # Add delay to respect server limits
  Sys.sleep(1)
  
  tryCatch({
    # Build the URL
    url <- paste0("https://portal.issn.org/resource/ISSN/", issn)
    
    # Make request
    response <- GET(url)
    
    if (status_code(response) == 200) {
      # Get the page content as text
      content_text <- content(response, "text", encoding = "UTF-8")
      
      # Find all ISSNs in the text
      all_issns <- unique(str_extract_all(content_text, "\\d{4}-\\d{3}[\\dX]")[[1]])
      
      # Filter out the input ISSN
      other_issns <- all_issns[all_issns != issn]
      
      # If there's another ISSN, it's likely the electronic one
      if (length(other_issns) > 0) {
        return(other_issns[1])
      }
    }
    
    return(NA_character_)
  }, error = function(e) {
    # Silently handle errors
    return(NA_character_)
  })
}

# First convert the electronic_issn column from logical to character
journals_crossref <- journals_crossref %>%
  mutate(electronic_issn = as.character(NA))

# Try with issn first
journals_with_eissn <- journals_crossref %>%
  mutate(electronic_issn = sapply(issn, get_eissn_simple))

# For rows where electronic_issn is still NA, try with the print_issn column
journals_with_eissn <- journals_with_eissn %>%
  mutate(electronic_issn = ifelse(
    is.na(electronic_issn) & !is.na(print_issn), 
    sapply(print_issn[is.na(electronic_issn) & !is.na(print_issn)], get_eissn_simple),
    electronic_issn
  ))
```


```{r}
# Save as excel file to resolve some remaining collapsed rows and errors
write_xlsx(journals_with_eissn, path = file.path(tempfolder, "OSHLINE_journals_eissn.xlsx")) 
```

Discordant ISSN from the original OSHLINE dataset and crossref were manually removed and checked against the ISSN portal. One journal was removed because its ISSN could not be confirmed in the ISSN portal (*Occupational Health and Safety (Tx.)*). 

```{r}
#| echo: true 
journals_with_eissn <- read_excel(path = file.path(psfolder,"OSHLINE_journals_eissn.xlsx")) 
```

```{r}
get_doaj_match <- function(issn, electronic_issn, print_issn) {
  
  Sys.sleep(0.5)
  match_found <- function(query_string, type = "issn") {
    # Skip if NA or empty
    if (is.na(query_string) || query_string == "") {
      return(FALSE)
    }
    
    base_url <- paste0("https://doaj.org/api/v2/search/journals/issn:", query_string)
    res <- GET(base_url)
    if (status_code(res) == 200) {
      json <- fromJSON(content(res, as = "text", encoding = "UTF-8"))
      return(length(json$results) > 0)
    }
    return(FALSE)
  }
  
  # Search in the specified order: issn, electronic_issn, print_issn
  return(any(
    match_found(issn),
    !is.na(electronic_issn) && match_found(electronic_issn),
    !is.na(print_issn) && match_found(print_issn)
  ))
}

journals_doaj <- journals_with_eissn %>%
  select(-issn_print_exact_match) %>% 
  rowwise() %>%
  mutate(doaj_match = get_doaj_match(issn, electronic_issn, print_issn)) %>%
  ungroup()
```

```{r}
write_xlsx(journals_doaj, path = file.path(tempfolder, "OSHLINE_journals_DOAJ_match.xlsx")) 
```

```{r}
#| echo: true 
journals_doaj <- read_excel(path = file.path(psfolder,"OSHLINE_journals_DOAJ_match.xlsx")) 
```

```{r}
journals_doaj %>% group_by(doaj_match) %>% 
  summarize(n = n()) %>% 
  mutate(percentage = round((n/sum(n) * 100),1))
```


