---
title: "Abstract EPICOH"
subtitle: "Open access indexing and publication costs of occupational health journals"
author: 
  - name: Javier Mancilla Galindo
    affiliation: Institute for Risk Assessment Sciences, Utrecht University, Utrecht, The Netherlands
    orcid: 0000-0002-0718-467X
    email: j.mancillagalindo@uu.nl
  - name: Antonio d’Errico
    affiliation: Institute for Risk Assessment Sciences, Utrecht University, Utrecht, The Netherlands
    orcid: 0000-0001-9805-600X
    email: a.derrico2@uu.nl
  - name: Sewon Lee
    affiliation: UCD Centre for Safety and Health at Work, School of Public Health, Physiotherapy, and Sports Science, University College Dublin, Dublin, Ireland
    orcid: 0000-0003-0618-1490
    email: sewon.lee@ucd.ie   
  - name: Max J. Oosterwegel
    affiliation: Institute for Risk Assessment Sciences, Utrecht University, Utrecht, The Netherlands
    orcid: 0000-0003-3991-2632
    email: m.j.oosterwegel@uu.nl
date: today
abstract: |
  <b>Objective:</b> Open access (OA) publishing allows research to be freely available to readers, whilst costs are usually transferred to authors, funders, or institutions. This study explores OA indexing and publication costs among occupational health journals.
  
  <b>Material and Methods:</b> We queried the Ulrichsweb database (a comprehensive catalog for academic and non-academic publications) for publications under the subject “occupational health and safety”. Filters were applied to restrict the search to active peer-reviewed journals with print and online formats. After deduplication, journals were searched for indexing in the Directory of Open Access Journals (DOAJ). Data on OA status, article processing charges (APC), and author copyright retention were retrieved from DOAJ. The DIAMAS criteria (persistent identification, scholarly journal, OA with licenses, no fees, open to all authors, and community-owned) were used to define diamond OA journals when all criteria were met. Publishers were labelled as commercial or non-commercial by reviewing publishers’ websites and stock exchange listings. APC were compared by type of publisher (commercial/non-commercial) with the Mann-Whitney U-test. A p < 0.05 was considered statistically significant.
  
  <b>Results:</b> Out of the 107 journals analyzed, 33 (30.8%) were fully OA and 15 (14%) were diamond OA journals. In the subset of fully OA journals, 6 (18.2%) were owned by commercial publishers and 27 (81.8%), non-commercial. Journals owned by commercial publishers charged higher APC (median: 1532.5 EUR, IQR: 671.25 – 1745) than non-commercial (median: 0 EUR, IQR: 0 – 421, p = 0.02). Fifteen (14%) OA journals allowed authors to retain copyright of their works. 
  
  <b>Conclusion:</b> Nearly one third of occupational health journals in Ulrichsweb are fully OA and less than 2 in 10, diamond OA. Those owned by commercial publishers charge higher APC. Existing occupational health journals could evaluate transitioning to non-commercial or diamond OA models, while new diamond OA journal initiatives could be encouraged.  
keywords: ["Occupational Health", "Open Access Publishing", "Bibliometrics", "Scholarly Communication", "Publishing"]
execute: 
  echo: false
  warning: false
format:
  html:
    toc: true
    toc_float: true
    embed-resources: true
    echo: true
    code-fold: true
    code-overflow: wrap
    code-tools: true
    code-links:
        - text: "GitHub"
          href: https://github.com/UtrechtUniversity/DOA-journal-occupational
          icon: github
  pdf: 
    toc: false
    documentclass: scrartcl
  docx: 
    toc: false
editor: source
---

```{r}
#| include: false  

# Create directories for sub-folders  
inputfolder <- "../data/raw"
psfolder <- "../data/processed"
tempfolder <- "../data/temp"
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"

dir.create(inputfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
dir.create(tempfolder, showWarnings = FALSE)
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
```

{{< pagebreak >}}

# Description 

```{r}
# Packages used in this session. 

if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  writexl,
  readxl,
  rvest,
  httr,
  jsonlite,
  gt,
  report
)

```


# Ulrichsweb

The [Ulrichsweb database](https://ulrichsweb.serialssolutions.com/) (Ulrichsweb.com™, Copyright © 2025 ProQuest LLC) was searched for publications under the subject "occupational health and safety" (code: 003980) on 07/05/2025. 

A total of **1418 records** were retrieved. The filter "Journals" was applied, resulting in **382 publications**. Subsequently, a filter for "Active" publications was applied, which led to **324 results**. A filter for "Refereed / Peer-reviewed" publications was subsequently applied, resulting in **205 publications**. Lastly, a filter for Print and Online publications was applied, resulting in 
**187 unduplicated records**. 

```{r}
# Load results of search in Ulrichsweb and clean dataset
source("scripts/ulrichsweb_data_cleaning.R")
```

Records were deduplicated by name as there was one entry for "Print" and another one for "Electronic" for journals available in both formats. The total number of records after deduplication was **n = `r count(journals_dedup)`**.   

```{r}
id_open_access <- ulrichsweb %>% filter(
  str_detect(Key.Features, "Open Access") 
) %>% pull(id) %>% unique
```

According to Ulrichsweb data, **n = `r length(id_open_access)` (`r (length(id_open_access)/count(journals_dedup) * 100) %>% round(1)`%)** of the journals are open access. This will later be compared against the number of matches in the DOAJ. 

```{r}
# Show the names of the journals that are open access
journals_dedup %>%
  filter(id %in% id_open_access) %>%
  distinct(id, Title) %>% 
  gt %>% 
  cols_label(id = "ID", Title = "Journal") %>% 
  tab_header(
    title = "Open Access (OA) journals in Ulrichsweb"
  )
```

{{< pagebreak >}}

# DOAJ 

All journals were searched in the Directory of Open Access Journals (DOAJ). 

```{r}
#| eval: false

source("scripts/DOAJ_match.R")
```

```{r}
#| include: false
journals_doaj <- read_excel(path = file.path(psfolder,"journals_DOAJ.xlsx")) 
```

```{r}
# id of journals with a doaj match

id_journals_doaj <- journals_doaj %>%
  filter(doaj_match == TRUE) %>%
  distinct(id) %>% pull

# Add column to journals_dedup indicating whether a journal was indexed in DOAJ
journals_dedup <- journals_dedup %>% 
  mutate(
    doaj_match = case_when(
      id %in% id_journals_doaj ~ TRUE, 
      T ~ FALSE
    )
  ) 
```

Out of the **n = `r count(journals_dedup)`** journals, a total of **n = `r length(id_journals_doaj)` (`r (length(id_journals_doaj)/count(journals_dedup) * 100) %>% round(1)`%)** were indexed in the Directory of Open Access Journals (DOAJ):   

```{r}
# Show the names of the journals that had a DOAJ match
journals_dedup %>%
  filter(doaj_match == TRUE) %>%
  distinct(id, Title) %>% 
  gt %>% 
  cols_label(id = "ID", Title = "Journal") %>% 
  tab_header(
    title = "Journals from Ulrichsweb indexed in DOAJ"
  )
```

```{r}
# Show differences between ids in id_journals_doaj and id_open_access
journals_doaj %>% 
  filter(
    id %in% id_journals_doaj & !id %in% id_open_access
  ) %>% 
  distinct(id, Title) %>% 
  gt() %>% 
  cols_label(id = "ID", Title = "Journal") %>% 
  tab_header(
    title = "Journals indexed in DOAJ but not listed as OA in Ulrichsweb"
  )
```

```{r}
# Show differences between ids in id_journals_doaj and id_open_access
journals_doaj %>% 
  filter(
    id %in% id_open_access & !id %in% id_journals_doaj
  ) %>% 
  distinct(id, Title) %>% 
  gt() %>% 
  cols_label(id = "ID", Title = "Journal") %>% 
  tab_header(
    title = "Journals listed as OA in Ulrichsweb but not indexed in DOAJ"
  )
```

```{r}
journals_analysis %>% filter(commercial_publisher == "yes") %>% 
  count()
```


A new column `open_access` was added to the `journals_dedup` dataframe, which indicates whether a journal was is OA either in Ulrichsweb or DOAJ.

```{r}
journals_dedup <- journals_dedup %>% 
  mutate(
    open_access = case_when(
      id %in% id_open_access ~ "yes", 
      id %in% id_journals_doaj ~ "yes",
      T ~ "no"
    )
  ) 
```

The status of the publisher as a *commercial* or *non-commercial* publisher was classified by reviewing the publishers' websites when available looking for declarations of for-profit opperations, or by reviewing listings of companies on a stock exchange (if true, these were labelled as commercial).

```{r}
# Create column of major publishers. Smaller entities will be manually added
journals_dedup <- journals_dedup %>% 
  mutate(
    commercial_publisher = case_when(
      # string detect out of multiple possible strings 
      str_detect(publisher, "Elsevier|Springer|BMJ|Biomed|Wiley|Taylor & Francis|Lippincott|Sage Publications|Routledge|Emerald Publishing|Wolters Kluwer") ~ "yes",
      str_detect(publisher, "Oxford University Press") ~ "no",
      T ~ NA_character_
    )
  ) 
```


```{r}
#| include: false
write_xlsx(journals_dedup, path = file.path(tempfolder, "journals_analysis.xlsx")) 
```

{{< pagebreak >}}

# Analysis of journals included. 

```{r}
#| include: false
journals_analysis <- read_excel(path = file.path(psfolder,"journals_analysis.xlsx")) 
```

## Open Access 

A total of **n = `r count(journals_analysis %>% filter(open_access == "yes"))` (`r (count(journals_analysis %>% filter(open_access == "yes"))/count(journals_analysis) * 100) %>% round(1)`%)** journals were fully open access (OA). 

## Diamond Open Access 

Journals were binary classified as diamond open access (DOA) and non-DOA based on meeting all six criteria stated in the [DIAMAS classification](https://diamasproject.eu/operational-diamond-oa-criteria-for-journals/). A total of **n = `r count(journals_analysis %>% filter(diamond_OA == "yes"))` (`r (count(journals_analysis %>% filter(diamond_OA == "yes"))/count(journals_analysis) * 100) %>% round(1)`%)** journals met all criteria to be classified as DOA:

```{r}
# Show the names of the journals that had a DOAJ match
journals_analysis %>%
  filter(diamond_OA == "yes") %>%
  distinct(id, Title, country) %>% 
  arrange(country) %>%
  gt %>% 
  cols_label(id = "ID", Title = "Journal", country = "Country")
```

## Article Processing Charges

Whenever APC charges were registered in a currency distinct to euros (EUR), values were converted to EUR by using international currency exchange rates from [Google Finance](https://www.google.com/finance/) on 07/05/2025.

The summary of journal article processing charges (APC) per article in euros (EUR) is shown in this table: 
```{r}
journals_analysis %>% 
  filter(open_access == "yes") %>% 
  summarize(
    n = n(),
    min = min(APC),
    Q1 = quantile(APC,0.25),
    median = median(APC),
    Q3 = quantile(APC,0.75),
    max = max(APC)
  ) %>% 
  gt %>% 
  cols_label(n = "Number of Journals") %>% 
  cols_align(
    align = "center",
    columns = everything()
    )
```

The summary of APC in EUR according to commercial status of the publisher is as follows: 

```{r}
journals_analysis %>% 
  filter(open_access == "yes") %>% 
  group_by(commercial_publisher) %>% 
  summarize(
    n = n(),
    min = min(APC),
    Q1 = quantile(APC,0.25),
    median = median(APC),
    Q3 = quantile(APC,0.75),
    max = max(APC)
  ) %>% 
  gt %>% 
  cols_label(
    n = "Number of Journals", 
    commercial_publisher = "Commercial Publisher"
    ) %>% 
  cols_align(
    align = "center",
    columns = everything()
    )
```

```{r}
boxplot(APC ~ commercial_publisher,
        data = journals_analysis %>% filter(open_access == "yes"),
        horizontal = TRUE,
        ylab = "Commercial publisher",
        xlab = "Article processing charges (APC) in EUR",
        col = "darkblue"
        )
```

The Mann–Whitney U-test (Wilcoxon's rank sum test) was applied to test the alternative hypothesis that the APC differed according to publisher commercial status classification: 

```{r}
wil <- wilcox.test(APC ~ commercial_publisher, 
                   data = journals_analysis, 
                   exact = FALSE
                   )
wil
```

## Year of conversion to OA 

The year of registration of journals as fully open access (OA) is summarized in the following table: 
```{r}
journals_analysis %>% 
  filter(open_access == "yes") %>%
  summarize(
    n = n(),
    min = min(OA_since),
    Q1 = quantile(OA_since,0.25),
    median = median(OA_since),
    Q3 = quantile(OA_since,0.75),
    max = max(OA_since)
  ) %>% 
  gt %>% 
  cols_label(n = "Number of Journals") %>% 
  cols_align(
    align = "center",
    columns = everything()
    )
```

## Author copyright  

Out of the `r count(journals_analysis)` publications included for analysis, **n = `r count(journals_analysis %>% filter(author_copyright == "yes"))` (`r ((count(journals_analysis %>% filter(author_copyright == "yes"))/count(journals_analysis)) * 100) %>% round(1)`%)** allowed authors to retain the copyright of their works. 

## Licensing 

The types of licenses allowed in the journals indexed in DOAJ were the following: 

```{r}
journals_analysis %>% 
  filter(open_access == "yes") %>%
  mutate(N = length(id)) %>% 
  separate_rows(license, sep = ", ") %>% 
  group_by(license) %>% 
  summarize(n = n(), N = first(N)) %>%
  arrange(desc(n)) %>% 
  mutate(Percentage = round((n/N * 100),1)) %>% 
  gt %>% 
  cols_hide(N)
```


{{< pagebreak >}}

# R Package References

```{r}
#| include: false  

# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL

session
```

```{r}
#| output: asis
report::cite_packages(session)
```

